---
title: "RNA-seq Differential Expression"
author: "Vaibhav"
date: "2025-12-20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,      # show code
  message = FALSE,  # hide package messages
  warning = FALSE,  # hide warnings
  fig.width = 7,
  fig.height = 5
)

```
---

**Introduction**

This analysis investigates hypoxia-induced transcriptional changes in prostate cancer
cell lines LNCaP (androgen-sensitive) and PC3 (androgen-independent).
Hypoxia is a major driver of tumor aggressiveness and lineage plasticity.

The goal is to identify:

[1]Differentially expressed genes (DEGs) under hypoxia

[2]Global expression patterns using PCA

[3]Hypoxia-regulated biological pathways

[4]Hallmark biological programs altered in LNCaP cells




**Loading required libraries**

These packages implement:

Differential expression statistics (DESeq2)

Data manipulation (tidyverse)

Visualization (ggplot2, pheatmap)

Pathway analysis (clusterProfiler, ReactomePA, fgsea)

***Packages must be installed before knitting***

```{r}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(matrixStats)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(fgsea)
```

**Loading raw count matrix**


```{r}
# Read count matrix

counts <- read.csv("GSE106305_counts_matrix.csv",
row.names = "Geneid",
stringsAsFactors = FALSE)


counts <- counts[, sort(colnames(counts))]
# View the first few rows
head(counts)

```


**Creating sample metadata**
```{r}

condition <- factor(c(
rep("LNCAP_Hypoxia", 2),
rep("LNCAP_Normoxia", 2),
rep("PC3_Hypoxia", 2),
rep("PC3_Normoxia", 2)
))


colData <- data.frame(condition)
rownames(colData) <- colnames(counts)
head(colData)
```

**DESeq2 dataset**

```{r}

dds <- DESeqDataSetFromMatrix(
countData = counts,
colData = colData,
design = ~ condition
)

# Keep genes with at least 10 reads in at least 2 samples
keep <- rowSums(counts(dds) >= 10) >= 2
dds <- dds[keep, ]
```


**Differential expression analysis**
```{r}
dds <- DESeq(dds)

#Normalized counts
norm_counts <- counts(dds, normalized = TRUE)
write.csv(norm_counts, "Normalized_counts.csv")

```

**Exploratory analysis: PCA**
```{r}

#Variance-stabilizing transformation
vsd <- vst(dds, blind = TRUE)

pca_data <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))


p_pca <- ggplot(pca_data, aes(PC1, PC2, color = condition)) +
geom_point(size = 3) +
labs(x = paste0("PC1: ", percentVar[1], "%"),
y = paste0("PC2: ", percentVar[2], "%"),
title = "PCA (VST transformed)") +
theme_minimal()


ggsave("PCA_plot.png", p_pca, width = 6, height = 5, dpi = 300)

print(p_pca)
dev.off()
```


**Sample-to-sample distance heatmap**
```{r}
sample_dists <- dist(t(assay(vsd)))
dist_matrix <- as.matrix(sample_dists)


png("sample_distance_heatmap.png", width = 1200, height = 1000, res = 300)
p <- pheatmap(dist_matrix,
clustering_distance_rows = sample_dists,
clustering_distance_cols = sample_dists,
col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
main = "Sample-to-sample distances")

print(p)
dev.off()
```


**Expression distribution diagnostics**
```{r}
png("density_raw_vs_vst.png", width = 2000, height = 2000, res = 300)
par(mfrow = c(2, 2))


plot(density(counts(dds)[,1]), main = "Raw counts", xlab = "Expression")
plot(density(assay(vsd)[,1]), main = "VST counts", xlab = "Expression")


dev.off()

```


**Highly variable gene heatmap**
```{r}
rv <- rowVars(assay(vsd))
top_genes <- order(rv, decreasing = TRUE)[1:40]


mat <- assay(vsd)[top_genes, ]
mat <- mat - rowMeans(mat)


png("top_variable_genes_heatmap.png", width = 1200, height = 1200, res = 300)
pheatmap(mat,
color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(255),
fontsize_row = 6,
main = "Top 40 most variable genes")
dev.off()
```
```{r}
dds_lncap <- dds[, grepl("LNCAP", colnames(dds))]
dds_lncap$condition <- droplevels(dds_lncap$condition)
dds_lncap$condition <- relevel(dds_lncap$condition, ref = "LNCAP_Normoxia")

```


**Extract differential expression results**
```{r}
dds_lncap <- DESeq(dds_lncap)


res_lncap <- results(dds_lncap,
contrast = c("condition",
"LNCAP_Hypoxia",
"LNCAP_Normoxia"))
write.csv(as.data.frame(res_lncap), "DEGs_LNCAP.csv")
head(res_lncap)

```

**MA plot**

```{r}

png("MA_plot_LNCAP.png", width = 800, height = 600, res = 150)
plotMA(res_lncap, ylim = c(-5, 5))
dev.off()
```


**Volcano plot**
```{r}

res_df <- as.data.frame(res_lncap) %>%
na.omit() %>%
mutate(regulation = case_when(
padj < 0.05 & log2FoldChange > 1 ~ "Upregulated",
padj < 0.05 & log2FoldChange < -1 ~ "Downregulated",
TRUE ~ "Not significant"
))


p_volcano <- ggplot(res_df, aes(log2FoldChange, -log10(padj), color = regulation)) +
geom_point(alpha = 0.6) +
theme_minimal() +
labs(title = "Volcano plot: LNCaP hypoxia")


ggsave("Volcano_plot_LNCAP.png", p_volcano, width = 6, height = 5, dpi = 300)
```


***Pathway Analysis using GSEA***



```{r}
#Converting gene IDs (ENSEMBL → ENTREZ)

res_lncap_df <- as.data.frame(res_lncap)
res_lncap_df$ENSEMBL <- rownames(res_lncap)

id_map <- bitr(
res_lncap_df$ENSEMBL,
fromType = "ENSEMBL",
toType = "ENTREZID",
OrgDb = org.Hs.eg.db
)

head(res_lncap_df)

#Merge mapping and removing duplicates
res_mapped <- res_lncap_df %>% left_join(id_map, by = "ENSEMBL") %>%
filter(!is.na(ENTREZID)) %>%
distinct(ENTREZID, .keep_all = TRUE)
head(res_mapped)
```


**Creating ranked gene list**

```{r}
gene_ranking <- res_mapped$log2FoldChange
names(gene_ranking) <- res_mapped$ENTREZID
gene_ranking <- sort(gene_ranking, decreasing = TRUE)

head(gene_ranking)
```

**Run GSEA using Reactome**
```{r}
gsea_reactome <- gsePathway(
geneList = gene_ranking,
organism = "human",
pvalueCutoff = 0.05,
verbose = FALSE
)

head(gsea_reactome@result)
```


***Over-Representation Analysis (ORA)***

```{r}
#extracted significant genes
sig_genes <- res_mapped %>%
filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
pull(ENTREZID)
```


**Run ORA using Reactome**
```{r}
ora_reactome <- enrichPathway(
gene = sig_genes,
organism = "human",
pvalueCutoff = 0.05
)

```

**Dotpot**
```{r}
d <-dotplot(ora_reactome, showCategory = 20)

ggsave("ora_reactome_dotplot.png", plot = d, width = 10, height = 8, dpi = 300)

d
```

***fgsea + Hallmark***

**Loaded fgsea and Hallmark gene sets**
```{r}

library(fgsea)


hallmark_sets <- gmtPathways("h.all.v7.0.symbols.gmt.txt")


# Convert ENSEMBL to SYMBOL
symbol_map <- bitr(
res_lncap$ENSEMBL,
fromType = "ENSEMBL",
toType = "SYMBOL",
OrgDb = org.Hs.eg.db
)


res_symbol <- res_lncap_df %>%
left_join(symbol_map, by = c("ENSEMBL" = "ENSEMBL")) %>%
filter(!is.na(SYMBOL)) %>%
distinct(SYMBOL, .keep_all = TRUE)


ranked_symbols <- res_symbol$log2FoldChange
names(ranked_symbols) <- res_symbol$SYMBOL


ranked_symbols <- sort(ranked_symbols, decreasing = TRUE)
```


**Run fgsea**
```{r}
fgsea_res <- fgsea(
pathways = hallmark_sets,
stats = ranked_symbols,
minSize = 15,
maxSize = 500,
nperm = 1000
)


fgsea_res[order(padj), .(pathway, NES, padj)][1:6]

```

**Visualized Hallmark enrichment**
```{r}

waterfall_plot <- function(fgsea_res, graph_title) {

  fgsea_df <- as.data.frame(fgsea_res)

  p <- fgsea_df %>%
    mutate(short_name = str_split_fixed(pathway, "_", 2)[,2]) %>%
    ggplot(aes(x = reorder(short_name, NES), y = NES)) +
      geom_bar(stat = "identity", aes(fill = padj < 0.05)) +
      coord_flip() +
      labs(
        x = "Hallmark Pathway",
        y = "Normalized Enrichment Score",
        title = graph_title
      ) +
      theme(
        axis.text.y = element_text(size = 7),
        plot.title = element_text(hjust = 0.5)
      )

  return(p)
}


p <- waterfall_plot(
  fgsea_res,
  "Hallmark pathways altered by hypoxia in LNCaP cells"
)

ggsave(
  filename = "Hallmark_fgsea_waterfall_LNCaP.png",
  plot = p,
  width = 8,
  height = 6,
  dpi = 300
)

```


